---
title: "hintreg:
  An R package for fitting a heterogeneous normal interval regression model
  to interval data with an indifference limen"
abstract: "Interval data with an indifference limen arise
  when a survey asks about a *change* of a continuous variable
  and the respondents choose from a set of given intervals including `no change'.
  This vignette introduces **hintreg**,
  an R package for fitting a heterogeneous normal interval regression model to such data.
  The model allows for heterogeneity not only in the location and scale parameters,
  but also in the lower and upper thresholds of the indifference limen,
  thus reducing the risk of model misspecification.
  An analysis of survey data on inflation expectations illustrates how to use the package."
bibliography: hintreg.bib
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{hintreg:
  An R package for fitting a heterogeneous normal interval regression model
  to interval data with an indifference limen}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

Survey questions about a continuous variable (e.g., income)
often ask respondents to choose from a set of given intervals.
Moreover,
for questions about a *change* of a continuous variable
(e.g., inflation, which is a change in the price level),
the choice may include a category `no change'.
This category does not literally mean a 0\% change,
whose probability is 0 for a continuous random variable,
but implies an *indifference limen*
that allows for an unnoticeable small change.
With known thresholds except for an indifference limen,
interval data with an indifference limen
differ from both interval and ordinal data.
Standard econometric softwares such as gretl and Stata
do not have commands to analyze such data easily.

Conceptually,
fitting a regression model to interval data with an indifference limen
is straightforward.
Just write down the likelihood function
and maximize it with respect to the parameters.
See @Murasawa13 for an example.^[
@Murasawa20 applies a Bayesian method,
assuming a normal mixture model.
]
One can also use an R package **oglmx** by @Carroll18
to fit an ordered response model with some known thresholds
(=interval regression model with some unknown thresholds),
even allowing for conditional heteroskedasticity.
However,
**oglmx** does not allow for heterogenous thresholds.^[
Some R packages for fitting an ordered response model
allow for heterogenous thresholds;
e.g., **hopit** by @Danko19 and **ordinal** by @Christensen19.
Typically they do not allow for known thresholds;
i.e., they cannot fit an interval regression model.
]


```{r setup}
library(hintreg)
```

# References
