---
title: "**hintreg**:
  An R package for fitting a heterogeneous normal interval regression model
  to interval data with an indifference limen^[
  This package was developed during my research visit
  at the Department of International Business, National Chengchi University,
  in the summer of 2022.
  I thank Natsuki Arai and Biing-Shen Kuo for their hospitality and support,
  and Konan University for funding the visit.
  This work was also supported by JSPS KAKENHI Grant Number JP21K01434.
  ]"
abstract: "Interval data with an indifference limen arise
  when a survey asks about a *change* of a continuous variable
  and the respondents choose from a set of given intervals including `no change'.
  This vignette introduces **hintreg**,
  an R package for fitting a heterogeneous normal interval regression model to such data.
  The model allows for heterogeneity not only in the location and scale parameters,
  but also in the lower and upper thresholds of the indifference limen,
  thus reducing the risk of model misspecification.
  An analysis of survey data on inflation expectations illustrates how to use the package."
bibliography: hintreg.bib
output:
  rmarkdown::pdf_document:
    number_sections: yes
  rmarkdown::html_vignette:
    number_sections: yes
    toc: yes
papersize: a4
vignette: >
  An R package for fitting a heterogeneous normal interval regression model
  to interval data with an indifference limen}
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{**hintreg**:
  %\VignetteEngine{knitr::rmarkdown}
---

```{r, include = FALSE}
options(width = 100)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = '100%'
)
```

```{r setup, echo = FALSE}
library(ggplot2)
library(hintreg)
```

# Introduction

Survey questions about a continuous variable (e.g., income)
often ask respondents to choose from a set of given intervals.
Moreover,
for questions about a *change* of a continuous variable
(e.g., inflation, which is a change in the price level),
the choice may include a category `no change'.
This category does not literally mean a 0\% change,
whose probability is 0 for a continuous random variable,
but implies an *indifference limen*,
which allows for an unnoticeable change.
With known thresholds except for an indifference limen,
interval data with an indifference limen
differ from both interval and ordinal data.
Standard econometric software packages such as gretl and Stata
do not have commands to analyze such data easily.

Conceptually,
fitting a regression model to interval data with an indifference limen
is straightforward.
Just write down the likelihood function
and maximize it with respect to the parameters.
See @Murasawa13 for an example.^[
@Murasawa20 applies a Bayesian method,
assuming a normal mixture model.
]
One can also use an R package **oglmx** by @Carroll18
to fit an ordered response model with some known thresholds
(=interval regression model with some unknown thresholds),
even allowing for conditional heteroskedasticity.
However,
**oglmx** does not allow for heterogeneous thresholds.^[
Some R packages for fitting an ordered response model
allow for heterogeneous thresholds;
e.g., **hopit** by @Danko19 and **ordinal** by @Christensen19.
They typically do not allow for known thresholds, however;
i.e., they cannot fit an interval regression model.
]

This vignette introduces **hintreg**,
an R package for fitting a heterogeneous normal interval regression model
to interval data with an indifference limen.
The model allows for heterogeneity
not only in the location and scale parameters,
but also in the lower and upper thresholds of the indifference limen.
Allowing for heterogeneity in the indifference limen
is not only interesting by itself,
but also reduces the risk of model misspecification,
which is important
because for a nonlinear model such as an interval regression model,
ignoring any kind of heterogeneity leads to inconsistency of the ML estimator.

The vignette proceeds as follows.
Section 2 defines a heterogeneous interval regression model
for interval data with an indifference limen.
Section 3 explains how to install **hintreg**
and use its main function **hintreg()**.
Section 4 analyzes survey data on inflation expectations
to illustrate how to use **hintreg**.

# Heterogeneous interval regression model

## Normal interval regression model

Let $(\boldsymbol{y},\boldsymbol{X})$
be a $(1+k)$-variate random sample of size $n$,
where $y_i$ is an ordered factor with $J+1$ levels.
A *normal interval regression model* for $y_i$ given $\boldsymbol{x}_i$
is for $i=1,\dots,n$,
\begin{align*}
y_i
& :=\begin{cases}
0      & \text{if $c_0<y^*_i \le c_1$} \\
\vdots & \\
J      & \text{if $c_J<y^*_i \le c_{J+1}$} \\
\end{cases} \\
y^*_i|\boldsymbol{x}_i
& \sim \mathrm{N}\left(\boldsymbol{x}_i'\boldsymbol{\beta},\sigma^2\right)
\end{align*}
where we know $\{c_j\}$.
This model is standard.
One can use **intreg** command in gretl and Stata to fit this model.

## Conditional heteroskedasticity

A *generalized normal interval regression model*
assumes a generalized normal linear regression model
for $y^*_i$ given $\boldsymbol{x}_i$,
i.e., for $i=1,\dots,n$,
\[y^*_i|\boldsymbol{x}_i \sim \mathrm{N}(
\boldsymbol{x}_i'\boldsymbol{\beta},
\exp(2\boldsymbol{x}_i'\boldsymbol{\delta})
)\]
This model is also standard.^[
Following **intreg** command in Stata,
we model conditional heteroskedasticity
not in variance but in standard deviation.
]
One can use **intreg** command in Stata to fit this model.

## Indifference limen

Assume that for $i=1,\dots,n$,
\begin{align*}
y_i
& :=\begin{cases}
0      & \text{if $\gamma_0<y^*_i \le \gamma_1$} \\
\vdots & \\
J      & \text{if $\gamma_J<y^*_i \le \gamma_{J+1}$} \\
\end{cases} \\
y^*_i|\boldsymbol{x}_i
& \sim \mathrm{N}(
\boldsymbol{x}_i'\boldsymbol{\beta},
\exp(2\boldsymbol{x}_i'\boldsymbol{\delta})
)
\end{align*}
where $\gamma_0<\dots<\gamma_l<0<\gamma_u<\dots<\gamma_{J+1}$
and we know $\{\gamma_j\}$ except for $[\gamma_l,\gamma_u]$,
which is an indifference limen.
This model is not standard.
However,
one can use an R package **oglmx** to fit this model.

## Heterogeneous indifference limen

The indifference limen $[\gamma_l,\gamma_u]$ may differ across $i$.
Let for $i=1,\dots,n$,
\begin{align*}
\gamma_{l,i}
& :=\bar{\gamma}_l\Lambda(\boldsymbol{x}_i'\boldsymbol{\lambda}) \\
\gamma_{u,i}
& :=\bar{\gamma}_u\Lambda(\boldsymbol{x}_i'\boldsymbol{\upsilon})
\end{align*}
where $\bar{\gamma}_l \in [\gamma_{l-1},0)$
is a lower bound on $\{\gamma_{l,i}\}$,
$\bar{\gamma}_u \in (0,\gamma_{u+1}]$
is an upper bound on $\{\gamma_{u,i}\}$,
and $\Lambda(.)$ is the logistic function.
This model is not standard.
We name it a *heterogeneous normal interval regression model*.
One can use our R package **hintreg** to fit this model.

# **hintreg** package

## Installation

**hintreg** is currently available from the author's repository on Github.
To install a package from Github,
it is convenient to use **install_github()** in **devtools**.
So one can start using **hintreg** in the following four steps:

1. install **devtools** from CRAN
1. load and attach **devtools**
1. install **hintreg** from Github
1. load and attach **hintreg**

or simply by running the following code:

```{r, eval = FALSE}
install.packages("devtools")
library(devtools)
install_github("ystmmrsw/hintreg")
library(hintreg)
```

## Available functions

**hintreg** has only three functions:
**intreg()**, **gintreg()**, and **hintreg()**.
The fist two functions are not new.
Given known thresholds,
they fit normal and generalized normal interval regression models,
respectively.
Here we introduce our main function **hintreg()**.

**hintreg()** takes the following arguments:

```{r, eval = FALSE}
hintreg(
  location,
  scale,
  lower,
  upper,
  data,
  start,
  weights,
  threshbelow,
  threshabove,
  limenlb,
  limenub,
  ...
)
```

Among these arguments,
**data**, **start**, and **weights** are standard.
So we do not discuss them here.

The first four arguments are all formulas:

**location**
  : specifies the mean; e.g., **y ~ x**

**scale**
  : specifies the standard deviation; e.g., **~ x**

**lower**
  : specifies the lower threshold of the indifference limen; e.g., **~ x**

**upper**
  : specifies the upper threshold of the indifference limen; e.g., **~ x**

The left-hand side of the formula in **location**
is the dependent variable, which must be an ordered factor.
The left-hand sides of the other formulas are empty.

The last four arguments are parameters
for the thresholds and the indifference limen:

**threshbelow**
  : is a vector of known thresholds below the indifference limen

**threshabove**
  : is a vector of known thresholds above the indifference limen
  
**limenlb**
  : is a lower bound on the heterogeneous indifference limen

**limenub**
  : is an upper bound on the heterogeneous indifference limen

If the dependent variable has $J+1$ levels with $J$ thresholds,
then there must be $J-2$ elements
in **threshbelow** and **threshabove** (one of them can be null),
excluding the thresholds of the indifference limen.
If not null,
then the elements in **threshbelow** and **threshabove**
must be negative and positive, respectively,
both in the increasing order.
Identification requires that $J \ge 4$.
To maintain the order of thresholds,
**limenlb** must be no less than **max(threshbelow)** and negative,
and **limenub** must be positive
and no greater than **min(threshabove)**.

Internally,
\texttt{hintreg()} maximizes the log-likelihood function
using \texttt{optim()}.
One can specify the optional arguments for \texttt{optim()}
in \texttt{hintreg()},
if desired.

# An illustration

## Survey data on inflation expectations

**hintreg** has a data set **ias2009febmay**
taken from the Bank of England Inflation Attitudes Survey.^[
The data set of the Inflation Attitudes Survey is available
at <https://www.bankofengland.co.uk/statistics/research-datasets>.
]
This data set allows us to replicate a result
in @Blanchflower-MacCoille09,
who study the formation of public inflation expectations in the UK.

The Inflation Attitudes Survey is a quarterly survey
starting from February 2001,
and collects a quota sample of adults throughout the UK.
The quota is 4,000 in February, and 2,000 in May, August, and November.
The survey asks both inflation perceptions and expectations,
and the respondents choose from the following eight intervals
(excluding `No idea'):

1. Gone/Go down
1. Not changed/change
1. Up by 1\% or less
1. Up by 1\% but less than 2\%
1. Up by 2\% but less than 3\%
1. Up by 3\% but less than 4\%
1. Up by 4\% but less than 5\%
1. Up by 5\% or more

So we have interval data with an indifference limen (choice 2).

The survey dates and variables in **ias2009febmay** are just enough
to replicate Table 3, column (3) in @Blanchflower-MacCoille09.
One can check the variables in the data set by summarizing it:

```{r}
summary(ias2009febmay)
```

The variable **q2** is the respondent's answer
to the question about inflation expectation over the next 12 months.
One can visualize its distribution among the respondents using a bar chart:

```{r}
ggplot(ias2009febmay) +
  geom_bar(aes(x = q2)) +
  coord_flip()
```

## Normal interval regression model

@Blanchflower-MacCoille09 implicitly assume
that the indifference limen is $[-.5,.5]$,
and fit normal interval regression models
to interval data on inflation expectations.
Since **hintreg()** assumes unknown thresholds for the indifference limen,
we cannot use **hintreg()** here.
Using **intreg()** instead,
one can replicate Table 3, column (3) in @Blanchflower-MacCoille09
as follows:

```{r}
intreg(
  q2 ~ age + work + sex + educ + tenure,
  data       = ias2009febmay,
  thresholds = c(-.5, .5, 1:5),
  method     = "BFGS"
) |> summary()
```

We want to introduce conditional heteroskedasticity
and a heterogeneous indifference limen into this model.
However,
ML estimation of such a general model with many explanatory variables
takes too much time for an illustration.
So we include only **sex** as an explanatory variable below;
i.e., we measure a gender gap in inflation expectations
without controlling for other variables.^[
Of course,
interested readers can try including more explanatory variables by themselves.
]

```{r}
lout <- intreg(
  q2 ~ sex,
  data       = ias2009febmay,
  thresholds = c(-.5, .5, 1:5),
  method     = "BFGS"
)
summary(lout)
```

We find that for the UK in February and May 2009,
men on average had significantly lower inflation expectations than women
by about .2\%.

## Generalized normal interval regression model

Now we introduce conditional heteroskedasticity.
Since the likelihood function may no longer be unimodal,
one should choose the starting values for the parameters carefully,
though the default starting values may work.
Here we set the starting values as follows:

mean equation
  : use the estimated coefficients in a normal interval regression model

standard deviation equation
  : use the estimated log standard deviation of the error term
  in a normal interval regression model as the intercept;
  set the other coefficients to 0

One can implement this procedure as follows:

```{r}
vstart_beta  <- coef(lout)
ck           <- length(vstart_beta)
vstart_delta <- c(lout$logscale, numeric(ck - 1))
vstart       <- c(vstart_beta, vstart_delta)
```

Since **hintreg()** assumes unknown thresholds for the indifference limen,
we cannot use **hintreg()** here.
Using **gintreg()** instead,
we can fit a generalized normal interval regression model to interval data
as follows:

```{r}
lout <- gintreg(
  q2 ~ sex,
     ~ sex,
  data       = ias2009febmay,
  start      = vstart,
  thresholds = c(-.5, .5, 1:5),
  method     = "BFGS"
)
summary(lout)
```

We find that for the UK in February and May 2009,
the standard deviations of inflation expectations among men and women
were not significantly different.

## Estimating the indifference limen

@Blanchflower-MacCoille09 implicitly assume
that the indifference limen is $[\gamma_l,\gamma_u]:=[-.5,.5]$.
We can use **hintreg()** to estimate $[\gamma_l,\gamma_u]$
by restricting the threshold equations to have only intercepts.
Here we set the starting values as follows:

mean equation
  : use the estimated coefficients for the mean equation
  in a generalized normal interval regression model

standard deviation equation
  : use the estimated coefficients for the standard deviation equation
  in a generalized normal interval regression model

thresholds
  : set the intercepts to 0,
  meaning that the starting values for $\gamma_l$ and $\gamma_u$
  are $\bar{\gamma}_l/2$ and $\bar{\gamma}_h/2$, respectively

One can implement this procedure as follows:

```{r}
vstart_beta  <- lout$coefL
vstart_delta <- lout$coefS
vstart_gamma <- numeric(2)
vstart       <- c(vstart_beta, vstart_delta, vstart_gamma)
```

The vectors of known thresholds below and above the indifference limen
are null and $(1,\dots,5)$, respectively.
It is natural to set $\bar{\gamma}_h:=1$.
By symmetry, we set $\bar{\gamma}_l:=-1$,
though it can be any negative number.
We can fit a heterogeneous normal interval regression model
to interval data with an indifference limen as follows:

```{r}
lout <- hintreg(
  q2 ~ sex,
     ~ sex,
     ~ 1,
     ~ 1,
  data        = ias2009febmay,
  start       = vstart,
  threshbelow = NULL,
  threshabove = 1:5,
  limenlb     = -1,
  limenub     = 1,
  method      = "BFGS"
)
summary(lout)
```

We find the following (for the UK in February and May 2009):

1. The intercept for the lower threshold is large
but not significantly different from 0,
meaning that $\hat{\gamma}_l$ is close to $-1$
but we cannot reject $H_0:\gamma_l=-.5$.
1. The intercept for the upper threshold is significantly larger than 0,
meaning that we reject $H_0:\gamma_u=.5$ in favor of $H_1:\gamma_u>.5$.

Since $\hat{\gamma}_l \approx -1$,
setting $\bar{\gamma}_l:=-1$ may be restrictive.
So we set $\bar{\gamma}_l:=-2$ and reestimate the model.

```{r}
lout <- hintreg(
  q2 ~ sex,
     ~ sex,
     ~ 1,
     ~ 1,
  data        = ias2009febmay,
  start       = vstart,
  threshbelow = NULL,
  threshabove = 1:5,
  limenlb     = -2,
  limenub     = 1,
  method      = "BFGS"
)
summary(lout)
```

Now we find the following (for the UK in February and May 2009):

1. The intercept for the lower threshold is positive
but not significantly different from 0,
meaning that $\hat{\gamma}_l<-1$ but we cannot reject $H_0:\gamma_l=-1$.
1. The intercept for the upper threshold
is not significantly different from 0,
meaning that we can no longer reject $H_0:\gamma_u=.5$.

## Heterogeneous normal interval regression model

Finally,
we introduce heterogeneity into the indifference limen.
Here we set the starting values as follows:

mean equation
  : use the estimated coefficients for the mean equation
  in a generalized normal interval regression model with an indifference limen

standard deviation equation
  : use the estimated coefficients for the standard deviation equation
  in a generalized normal interval regression model with an indifference limen

threshold equations
  : use the estimated thresholds
  in a generalized normal interval regression model with an indifference limen
  as the intercepts;
  set the other coefficients to 0

One can implement this procedure as follows:

```{r}
vstart_beta  <- lout$coefL
vstart_delta <- lout$coefS
ck           <- length(vstart_beta)
vstart_lower <- c(lout$coefLower, numeric(ck - 1))
vstart_upper <- c(lout$coefUpper, numeric(ck - 1))
vstart       <- c(vstart_beta, vstart_delta, vstart_lower, vstart_upper)
```

Since $\hat{\gamma}_l<-1$ in the previous result,
we set $\bar{\gamma}_l:=-2$,
and fit a heterogeneous normal interval regression model
to interval data with an indifference limen as follows:

```{r}
hintreg(
  q2 ~ sex,
     ~ sex,
     ~ sex,
     ~ sex,
  data        = ias2009febmay,
  start       = vstart,
  threshbelow = NULL,
  threshabove = 1:5,
  limenlb     = -2,
  limenub     = 1,
  method      = "BFGS"
) |> summary()
```

We find the following (for the UK in February and May 2009):

1. The intercepts for the lower and upper thresholds
are significantly larger than 0,
meaning that we reject $H_0:\gamma_l=-1$ and $H_0:\gamma_u=.5$
in favor of $H_1:\gamma_l<-1$ and $H_1:\gamma_u>.5$, respectively,
at least for women.
2. The indifference limens for men and women
are not significantly different.

# References
